---
title: "140.654 Lab 1"
author: "Erjia Cui & Elizabeth Colantuoni"
output: html_document
---

The objectives of this lab session are (1) Review interpretation of regression coefficients from a logistic regression model; (2) Connect logistic regression coefficients to commonly computed statistics from 2x2 tables; (3) Use tables to display effect modification / interaction for binary outcomes and two binary covariates

```{r,message=FALSE,warning=FALSE}
library(ggplot2)
library(dplyr)
load('./nmes.rdata')
data <- nmes
data[data == '.'] <- NA
```

### National Medical Expenditure Survey (NMES) dataset

* Outcome: big expenditure defined as total expenditure > 1000 (bigexp)
* Exposure: Major smoking caused disease (mscd)
* Covariate: person's age (older: whether his/her age >= 65 yrs)
* Dataset: 11,684 people for 40 or older who have smoking data available

```{r}
data$posexp <- ifelse(data$totalexp > 0, 1, 0)
data$mscd <- ifelse(data$lc5+data$chd5 > 0, 1, 0)
data1 <- data[!is.na(data$eversmk),]

# total sample
data1$older <- ifelse(data1$lastage < 65, 0, 1)
data1$bigexp <- ifelse(data1$totalexp > 1000, 1, 0)
dim(data1)

# table (stratify by mscd, old)
mytable <- table(bigexp = data1$bigexp, mscd = data1$mscd, older = data1$older)
ftable(mytable) # flatten the contingency table

```

### 1. Model C: intercept + mscd + old + mscd:old

Suppose we want to understand the relationship between having a big expenditure and MSCD separately for younger and older persons. We first make contingency tables for each group.

__1. Table 1: Within younger strata, tabulate big expenditure and mscd__
```{r}
table(bigexp = data1$bigexp[data1$older == 0], mscd = data1$mscd[data1$older == 0])
```

__2. Table 2: Within older strata, tabulate big expenditure and mscd__
```{r}
table(bigexp = data1$bigexp[data1$older == 1], mscd = data1$mscd[data1$older == 1])
```

__3. Table 3: Within total sample, tabulate big expenditure and mscd__
```{r}
table(bigexp = data1$bigexp, mscd = data1$mscd)
```

The question of interest can be specified as the following model:

$$
Model C: logit [ Pr(Y = 1 | MSCD, Old) ] = \beta_0 + \beta_1\cdot MSCD + \beta_2 \cdot Old + \beta_3 \cdot MSCD \cdot Old
$$


Write down all possible scenarios:

* For younger population: \begin{equation} logit [ Pr(Y = 1 | MSCD, Old=0) ] = \beta_0  + \beta_1\cdot MSCD\end{equation}
* For older popultion: \begin{equation} logit [ Pr(Y = 1 | MSCD, Old=1) ] = (\beta_0 + \beta_2) + (\beta_1 + \beta_3) \cdot MSCD \end{equation}

__In class exercise: How to derive the estimates and SEs?__

Derive the estimate and SE for $\beta_0, \beta_1, \beta_0 + \beta_2, \beta_1 + \beta_3$ using Table1-3. Also, try to think about how to interpret these estimates?

</br></br></br>


* $\beta_0$: log odds of having big expenditure in the younger population without a MSCD (Refer to Table1)
\begin{align*}
\hat{\beta_0} &= log(\frac{1802}{4780}) = - 0.976\\
SE(\hat{\beta_0}) &= \sqrt{\frac{1}{4780} + \frac{1}{1802}} = 0.0276
\end{align*}

* $\beta_0 + \beta_2$: log odds of having big expenditure in the older population without a MSCD (Refer to Table2)

\begin{align*}
\hat{\beta_0}+\hat{ \beta_2} &= log(\frac{1547}{2236}) = -0.368\\
SE(\hat{\beta_0}+ \hat{\beta_2}) &= \sqrt{\frac{1}{1547} + \frac{1}{2236}} = 0.033
\end{align*}

* Then, $\beta_2$:
\begin{align*}
\hat{ \beta_2} &= \hat{\beta_0}+\hat{ \beta_2}-\hat{\beta_0} = -0.368 -(-0.976) = 0.608\\
SE(\hat{\beta_2}) &= \sqrt{SE(\hat{\beta_0}+ \hat{\beta_2})^2  + SE(\hat{\beta_0})^2} = 0.043
\end{align*}


* $\beta_1$:log odds ratio of having big expenditure comparing the younger persons with a MSCD to younger persons without a MSCD. An alternative interpretation is: the difference in the log odds of having a big expenditure for younger persons with a MSCD and the log odds of having a big expenditure for younger persons without a MSCD (Refer to Table1)

\begin{align*}
\hat{\beta_1} &= log(\frac{273/101}{1802/4780}) = 1.97\\
SE(\hat{\beta_1}) &= \sqrt{\frac{1}{273} +\frac{1}{101}+\frac{1}{1802}+ \frac{1}{4780}} = 0.1197
\end{align*}

* $\beta_1 + \beta_3$: log odds ratio of having big expenditure comparing the older persons with a MSCD to older persons without a MSCD. (Refer to Table2)

\begin{align*}
\hat{\beta_1}+\hat{ \beta_3} &= log(\frac{713/232}{1547/2236}) = 1.491\\
SE(\hat{\beta_1}+ \hat{\beta_3}) &= \sqrt{\frac{1}{713} +\frac{1}{232}+\frac{1}{1547}+ \frac{1}{2236}} = 0.0825
\end{align*}


* $\beta_3$: difference in the log odds ratio of having big expenditure comparing older persons with and without a MSCD and the log odds ratio of having a big expenditure comparing younger persons with and without a MSCD

For $\beta_3$: log of ratio of two OR
\begin{align*}
\beta_1 &= log[OR(MSCD,Old=0)] = log(\frac{Pr(Y|MSCD,Old=0)}{1 - Pr(Y|MSCD,Old=0)})\\
\beta_1 + \beta_3 &= log[OR(MSCD,Old=1)]\\
\beta_3 &= (\beta_1 + \beta_3)-\beta_1 = log[OR(MSCD,Old=1)] - log[OR(MSCD,Old=0)] = log(\frac{OR(MSCD,Old=1)}{OR(MSCD,Old=0)})
\end{align*}

\begin{align*}
\hat{ \beta_3} &= \hat{\beta_1}+\hat{ \beta_3}-\hat{\beta_1} = 1.491 -(1.97) = -0.479\\
SE(\hat{\beta_3}) &= \sqrt{SE(\hat{\beta_1}+\hat{ \beta_3})^2  + SE(\hat{\beta_1})^2} = 0.145
\end{align*}

__Using GLM in R and Compare with the above derivation__

Recall the derivation:
$$
\hat{\beta_0} = - 0.976; SE_{\hat{\beta_0}} = 0.0276\\
\hat{\beta_1} = 1.97; SE_{\hat{\beta_1}} = 0.1197\\
\hat{\beta_2} = 0.608; SE_{\hat{\beta_2}} = 0.043\\
\hat{\beta_3} = -0.479; SE_{\hat{\beta_3}} = 0.145\\
$$

```{r}
data1$mscd.older <- data1$mscd*data1$older
lrC <- glm(bigexp~mscd+older+mscd.older, data=data1, family=binomial(link="logit"))
summary.glm(lrC)
```

They are identical!

### 2. Revisit Logistic regression models (A, B)

__Model A: intercept only__

$$
Model A:  logit [Pr(Y = 1) ] = \beta_0
$$


```{r}
lrA <- glm(bigexp~1, data=data1, family=binomial(link="logit"))
summary.glm(lrA)
```

__Model B: intercept + mscd__

$$
Model B: logit [Pr(Y = 1 | MSCD )] = \beta_0 + \beta_1 \cdot MSCD
$$
```{r}
lrB <- glm(bigexp~mscd, data=data1, family=binomial(link="logit"))
summary.glm(lrB)
```

As you might observe, the estimates are the same as the ones we have derived in class.

